Spring haul-out behavior of seals in the Bering and Chukchi seas

### 2024-02-17 analysis/paper/London_HauloutBeahvior.Rmd
pconn revision comments addressed

The remoteness of the Bering and Chukchi seas means direct scientific
observation of seal behavior is impractical. Thus, bio-logging devices are
especially useful tools for collecting key information on movement and haul-out
behavior for these species. Bio-logging records of time spent out of the water
provide valuable data for identifying covariates that explain variation in
haul-out behavior. For instance,Von Duyke et al. [-@vonduyke2020] used
satellite-linked bio-loggers to corroborate seasonal changes between diurnal and
nocturnal haul-out behavior of ringed seals previously described by Kelly and
Quakenbush [-@kelly1990c] using VHF radio tags and direct observation. Bengtson
et al. [-@bengtson2005a] documented a higher propensity for ringed seal basking
near solar noon, as did Ver Hoef et al. [-@verhoef2014a] in an analysis of
bearded, ribbon, and spotted seals using much larger sample sizes. Olnes et al.
[-@olnes2020] showed that the proportion of time bearded seals spent hauled out
progressively increased through spring and summer, and Ver Hoef et al.
[-@verhoef2014a] found haul-out probabilities increased gradually starting in
March and peaked in May and June for bearded, ribbon, and spotted seals. Recent
Such
analyses have not been limited to the Arctic. In the Antarctic, Bengtson and
Cameron [-@bengtson2004] relied on bio-logging data to demonstrate greater
haul-out propensity in juvenile crabeater seals (_Lobodon carcinophaga_) than
adults, with highest probabilities in February and at times close to solar noon.

@@

Haul-out behavior data were recorded in a manner standard across Wildlife Computers
bio-loggers and transmitted via the Argos satellite
network as hourly percent-dry timelines. For each hour of a day, the wet/dry
sensor was polled by the tag firmware every few seconds and the percent of the
hour in the dry state was calculated (Figure \@ref(fig:examplePlot)). On board
the bio-logger, hourly percent-dry calculations were rounded to the nearest 10%
inclusive of 0% and 100% along with additional values at 3% and 98%. This
compression resulted in additional data transmission as each message consisted
of two complete 24-hour records. Memory capacity allowed caching of percent-dry
records for several weeks or months and each message was transmitted several
times to ensure reception at the satellite. Bio-loggers were, typically,
deployed and programmed in a manner to maximize data transmission during the
spring pupping and molting period, though hourly percent-dry data were not
always successfully transmitted. This is due to a variety of factors including
satellite coverage, tag availability (e.g., tags mounted to the rear flipper
often do not transmit while at sea), tag performance, duty cycling, and
atmospheric interference. Fortunately, missing records do not substantially bias
inference about haul-out probabilities [@conn2012c].

@@

In addition to sex and age class, we analyzed variables that might help explain
variation in haul-out probabilities. These included day-of-year (for seasonal
effects) and local solar hour (for diurnal effects). We calculated local solar
hour using the {solaR} package [@perpinan2012] within the R statistical
environment [@rcoreteam2021] based on the weighted daily average locations. We
also linked the weighted average daily locations to weather values from the
North American Regional Reanalysis (NARR) model produced by the National Centers
for Environmental Prediction [@mesinger2006a]. The NARR model assimilates
observational data to produce a long-term picture of weather over North America
and portions of the surrounding seas. Weather variables are made available
across the region 8 times daily. For this study, NARR weather values were subset
to the extent of our study area over the Bering and Chukchi seas at 3-hr
intervals based on the native grid resolution of 32 km (1024 km^2^). The
following meteorological variables are known to affect haul-out behavior in
other Arctic pinnipeds [@reder2003; @udevitz2009; @perry2017] and were
interpolated and assigned to daily seal locations using a bilinear method: 1)
air temperature at 2 m above the Earth's surface, 2) wind consisting of
northerly and easterly vector components converted to wind speed using the
Euclidean norm, 3) barometric pressure at sea level, and 4) precipitation 
(Table \@ref(tab:tblCovariates)).

@@

To properly account for temporal dependence and to take advantage of computational
efficiency, we used generalized linear mixed pseudo-models (GLMPMs;
[@verhoef2010a]) to model variation in haul-out behavior as a function of (1)
covariate predictors, (2) temporally autocorrelated random effects, and (3)
individual random effects representing heterogeneity in individual behavior. We
used the glmmLDTS package [@verhoef2010a] to implement GLMPMs. We explored two
different model formulations for our data and we fit separate models to bearded, 
ribbon, and spotted seal data sets as we expected differing behavior by species. 
Separate models for each species were also
needed because a single, very large data set proved computationally intractable.
In our first model formulation and for each species, we fitted a year-independent
model that predicted average haul-out behavior as a function of demographic,
environmental, seasonal, and diurnal effects. Second, for ribbon and spotted
seals (which had considerably more data than bearded seals), we fitted models
that included all the effects from the first model, but also permitted annual
variation in haul-out timing. This second class of models was used to examine
whether haul-out patterns varied by year and to determine the annual timing of
apparent peaks in haul-out behavior. For both models, we assumed an hourly
Bernoulli response (i.e., whether tags were mostly dry or mostly wet) where
the linear predictor was modeled on the logit scale. This is consistent with
previous approaches [@verhoef2014a, @london2012] and only
`r (1 - round((portion_10(analysis_data) + portion_90(analysis_data))) * 100)`% of our
observations fell between 10% and 90% hourly percent-dry.

@@

Figure \@ref(fig:dataMap) shows the spatial distribution of weighted locations
with available haul-out behavior data used for analysis of each species across
the study area. Figure \@ref(fig:dataCal) shows the temporal distribution of all
haul-out data across the study season for each species. Observations for ribbon
and spotted seals were concentrated in the months of May and June due to the
timing of deployment (April and May) and the timing of molt (May and June).
During molt, seals (and their attached bio-loggers) spend more time out of the
water and more data are transmitted. Molt timing also impacts when many
deployments end as any bio-loggers adhered to the hair will fall off. Relative
to the other species in the study, there were fewer deployments of bio-loggers
on bearded seals. This resulted in fewer data observations overall and
noticeably lower in numbers May and June. The majority of bearded seal
deployments started later in the summer and, by May, bio-loggers had either
fallen off or their batteries were depleted.

@@

Ribbon seals exhibited a pattern of gradually increasing haul-out probability
in April that peaks in late May for subadults and in early June for
adults (Figure \@ref(fig:ribbonHOCal); see also 
@\ref(fig:ribbonPredSE)). The behavior was clearly centered
around local solar noon and expanded to other hours later in the spring
as seals entered their molting period. Subadults showed an earlier start
and more intense haul-out activity in April and May. The
young-of-the-year records begin after weaning and the model predictions
demonstrated the ontogeny of in-water activities (e.g. swimming, foraging)
in May. Adult females had a more protracted haul-out season compared to
males, and more time was spent hauled out in June compared to adult males and
subadults.

@@

Unlike previous analyses of seal haul-out behavior in spotted and ribbon seals
(e.g. [@verhoef2010a], [@conn2014a]), we also investigated the influence of
sex-age class on haul-out probabilities for both species (not for bearded seals
because of low sample size). Field identification of age class can be inexact,
particularly when differentiating subadults from adults. In the case of ribbon
seals, subadults often have less distinct ribbons than adults. Spotted seal
pelage cannot be used to reliably discern adults from subadults. Despite these
challenges, we feel the age classifications used in this analysis are useful in
testing for age-related effects on haul-out behavior.

Although ribbon and spotted seals exhibited a
unimodal diel haul-out pattern generally centered around local solar noon, there
were key differences across species, age, and sex that match our understanding
from natural history descriptions of their ecological behavior. Spotted seals
are known to form triads during the breeding season where a female and dependent
pup are accompanied on the ice by a suitor male [@frost2018]. The male waits for
the female to wean the pup and enter estrus, and fends off any other potential
suitor males. Triad formation results in both males and females spending a large
portion of the day hauled out on ice and a protracted spring haul-out season for
both sexes. While females are still nursing the pup and not yet in estrus they
may be less inclined to interrupt their haul out and enter the water where the
suitor male would attempt mating. We see this reflected in the predicted
haul-out patterns, with both males and females exhibiting a broad distribution
of time out of the water throughout the solar day and the season. Ribbon seals
are not known to form triads and our model predicts a progression of increased
haul-out behavior with females starting earlier in the season than males.
Notably, female ribbon seals spend a large portion of the day in the water
during the pupping period, aligning with the hypothesis that ribbon seal females
continue foraging while nursing. Subadult ribbon and spotted seals begin
elevated haul-out behavior earlier in the spring and follow a pattern seen in
other phocids where yearlings and subadults molt first followed by adult females
and males [@reder2003, @thompson1987, @kirkman2003].

We also investigated the influence of weather on haul-out probabilities,
including wind speed, temperature, barometric pressure, precipitation, and wind
chill. These have been investigated for walruses (e.g. Udevitz et al.
[-@udevitz2009]) and a few select studies of ice-associated seals
[@hamilton2018, @perry2017]. In our study, ribbon seals seemed to be the most
influenced by weather, with wind, temperature, and barometric pressure all being
important components of the model. Spotted seals were most affected by wind and
barometric pressure. For bearded seals, the model indicated wind and
temperature had the greatest impact. In general, and as might be expected, seals
were more likely to haul-out when daily temperatures were warmer, winds speeds
were lower, barometric pressure was higher, and precipitation was lower. Those
weather conditions are general indicators of increased solar radiation and lower
convective heat loss, both of which provides energetic benefits (see additonal
discussion in Supplemental Material [S3][#s3] regarding the potential use of
solar radiation directly). Low winds and
precipitation could also enhance predator detection. Our results highlight the
importance of incorporating weather covariates when analyzing haul-out behavior
and calculating availability corrections for aerial surveys.

@@

Consider
the case of employing haul-out probabilities as an aerial survey correction in a
population inhabiting two areas: an ice-free region and a region with sea ice.
Further, assume that half of the population is in each area on average, and that
the probability of hauling out is 1.0 in the ice covered region and 0.0 in the
region without ice.  Denoting the total population as $N$, let us examine what
happens when we use (1) population level availability, and (2) ice-specific
availability as a correction factor.  We'll further assume, for sake of this
example, that detection probability of seals on ice is 1.0, and that the
entirety of the ice covered region is censused. In case (1), our aerial survey
count is $C = 0.5 \cdot N$, with a population-level availability probability of
$\hat{a} = 0.5$. A Horvitz-Thompson-type estimator for abundance is simply 
$\hat{N} = C \div \hat{a}$, which has expectation 
$\mathbb{E}(\hat{a}) = \frac{0.5 \cdot N}{0.5} = N$, as desired.  Now
consider case (2).  In this case, the probability of hauling out is 1.0 for
seals in the ice-covered area, so we have 
$\mathbb{E}(\hat{N}) = \frac{0.5 \cdot N}{1.0} = 0.5N$. That
is, we bias abundance because we are not accounting for seals that are away from
ice, and therefore have a zero chance of hauling out. The same logic holds for
any other covariate, such as distance from the ice edge, that has different
values in the surveyed and unsurveyed habitat of the seals (assuming distance
from ice edge is coded with opposite signs inside and outside the pack ice).

@@

Previous attempts to estimate the abundance of phocid seals from aerial survey
data in the Bering and Chukchi seas (e.g. @bengtson2005a, @conn2014a,
@verhoef2014a) have used estimated haul-out probabilities to correct for the
proportion of animals that are in the water and thus unavailable to be counted.
Although several of these studies allowed haul-out probabilities to vary by
day-of-year and time-of-day, they have not accounted for variability among
years, in weather conditions, and in the age-sex class of the sample. In this
paper, we have shown that there can be considerable differences in the number of
seals hauled out on ice based on these factors. We recommend that future
abundance analyses employ availability models that account for them. For
instance, it is relatively straightforward to obtain weather reanalysis products
for times and locations that are surveyed and to construct a relevant correction
factor based on predictions of GLMPMs. The most challenging element in
developing availability correction factors is with annual variability. It can be
difficult to get a sufficient sample size to estimate year-specific correction
factors, particularly because research teams would likely need to tag seals and
conduct aerial surveys concurrently, requiring considerably more personnel
and money. One possible suggestion is to estimate a "shift" parameter within
models for aerial survey counts that allow the peak of haul-out distributions to
be adjusted earlier or later in the year based on the frequency of counts
observed over time. Regardless, researchers should anticipate there being some
unmodeled heterogeneity in availability when using aerial surveys to estimate
Arctic seal abundance. This may require consideration in trend estimation,
as one will not know if moderate differences in abundance estimates are
attributable to changes in abundance or changes in haul-out behavior.

@@

We focused this paper on haul-out behavior of bearded, ribbon, and spotted
seals. Ringed seals are also present in the Bering and Chukchi seas but exhibit
a unique complicating factor. Adult ringed seals build subnivean lairs under the
snow on top of the sea ice, where they haul out and where females rear pups
until conditions are good for basking [@frost2004]. Thus, the wet-dry sensor on
a bio-logger could indicate that an animal is hauled out, but if it is within a
lair, it is not available to be detected during an aerial survey. We hope to
address availability of ringed seals using data from satellite tags, replicate
survey tracks, and auxiliary information about snow depth and timing of melt in
a future study.

>>>>

The remoteness of the Bering and Chukchi seas means direct scientific
observation of seal behavior is impractical. Thus, bio-logging devices are
especially useful tools for collecting key information on movement and haul-out
behavior for these species. Bio-logging records of time spent out of the water
provide valuable data for identifying covariates that explain variation in
haul-out behavior. For instance,Von Duyke et al. [-@vonduyke2020] used
satellite-linked bio-loggers to corroborate seasonal changes between diurnal and
nocturnal haul-out behavior of ringed seals previously described by Kelly and
Quakenbush [-@kelly1990c] using VHF radio tags and direct observation. Bengtson
et al. [-@bengtson2005a] documented a higher propensity for ringed seal basking
near solar noon, as did Ver Hoef et al. [-@verhoef2014a] in an analysis of
bearded, ribbon, and spotted seals using much larger sample sizes. Olnes et al.
[-@olnes2020] showed that the proportion of time bearded seals spent hauled out
progressively increased through spring and summer, and Ver Hoef et al.
[-@verhoef2014a] found haul-out probabilities increased gradually starting in
March and peaked in May and June for bearded, ribbon, and spotted seals. Such
analyses have not been limited to the Arctic. In the Antarctic, Bengtson and
Cameron [-@bengtson2004] relied on bio-logging data to demonstrate greater
haul-out propensity in juvenile crabeater seals (_Lobodon carcinophaga_) than
adults, with highest probabilities in February and at times close to solar noon.

@@

Haul-out behavior data were recorded in a manner standard across Wildlife
Computers bio-loggers and transmitted via the Argos satellite network as hourly
percent-dry timelines. For each hour of a day, the wet/dry sensor was polled by
the tag firmware every few seconds and the percent of the hour in the dry state
was calculated (Figure \@ref(fig:examplePlot)). On board the bio-logger, hourly
percent-dry calculations were rounded to the nearest 10% inclusive of 0% and
100% along with additional values at 3% and 98%. This compression resulted in
additional data transmission as each message consisted of two complete 24-hour
records. Memory capacity allowed caching of percent-dry records for several
weeks or months and each message was transmitted several times to ensure
reception at the satellite. Bio-loggers were deployed and programmed in a manner
to maximize data transmission during the spring pupping and molting period,
though hourly percent-dry data were not always successfully transmitted. This is
due to a variety of factors including satellite coverage, tag availability
(e.g., tags mounted to the rear flipper often do not transmit while at sea), tag
performance, duty cycling, and atmospheric interference. Fortunately, missing
records do not substantially bias inference about haul-out probabilities
[@conn2012c].

@@

In addition to sex and age class, we analyzed variables that might help explain
variation in haul-out probabilities. These included day-of-year (for seasonal
effects) and local solar hour (for diurnal effects). We calculated local solar
hour using the {solaR} package [@perpinan2012] within the R statistical
environment [@rcoreteam2021] based on the weighted daily average locations. We
also linked the weighted average daily locations to weather values from the
North American Regional Reanalysis (NARR) model produced by the National Centers
for Environmental Prediction [@mesinger2006a]. The NARR model assimilates
observational data to produce a long-term picture of weather over North America
and portions of the surrounding seas. Weather variables are made available
across the region 8 times daily. For this study, NARR weather values were subset
to the extent of our study area over the Bering, Chukchi, and Beaufort seas at 3-hr
intervals based on the native grid resolution of 32 km (1024 km^2^). The
following meteorological variables are known to affect haul-out behavior in
other Arctic pinnipeds [@reder2003; @udevitz2009; @perry2017] and were
interpolated and assigned to daily seal locations using a bilinear method: 1)
air temperature at 2 m above the Earth's surface, 2) wind consisting of
northerly and easterly vector components converted to wind speed using the
Euclidean norm, 3) barometric pressure at sea level, and 4) precipitation 
(Table \@ref(tab:tblCovariates)).

@@

To properly account for temporal dependence and to take advantage of computational
efficiency, we used generalized linear mixed pseudo-models (GLMPMs;
[@verhoef2010a]) to model variation in haul-out behavior as a function of (1)
covariate predictors, (2) temporally autocorrelated random effects, and (3)
individual random effects representing heterogeneity in individual behavior. We
used the glmmLDTS package [@verhoef2010a] to implement GLMPMs. We explored two
different model formulations for our data and we fit separate models to bearded, 
ribbon, and spotted seal data sets as we expected differing behavior by species. 
Separate models for each species were also
needed because a single, very large data set proved computationally intractable.
In our first model formulation and for each species, we fitted a year-independent
model that predicted average haul-out behavior as a function of demographic,
environmental, seasonal, and diurnal effects. Second, for ribbon and spotted
seals (which had considerably more data than bearded seals), we fitted models
that included all the effects from the first model, but also permitted annual
variation in haul-out timing. This second class of models was used to examine
whether haul-out patterns varied by year and to determine the annual timing of
apparent peaks in haul-out behavior. For both models, we assumed an hourly
Bernoulli response (i.e., whether tags were mostly dry or mostly wet) where
the linear predictor was modeled on the logit scale. This is consistent with
previous approaches [@verhoef2014a, @london2012] and only
`r (1 - (portion_10(analysis_data) + portion_90(analysis_data))) * 100`% of our
observations fell between 10% and 90% hourly percent-dry.

@@

Figure \@ref(fig:dataMap) shows the spatial distribution of weighted locations
with available haul-out behavior data used for analysis of each species across
the study area. Figure \@ref(fig:dataCal) shows the temporal distribution of all
haul-out data across the study season for each species. Observations of ribbon
and spotted seals were concentrated in the months of May and June due to the
timing of deployment (April and May) and the timing of molt (May and June).
During molt, seals (and their attached bio-loggers) spend more time out of the
water and more data are transmitted. Molt timing also impacts when many
deployments end as any bio-loggers adhered to the hair will fall off. Relative
to the other species in the study, there were fewer deployments of bio-loggers
on bearded seals. This resulted in fewer data observations overall and
noticeably lower in numbers May and June. The majority of bearded seal
deployments started later in the summer and, by May, bio-loggers had either
fallen off or their batteries were depleted.

@@

Ribbon seals exhibited a pattern of gradually increasing haul-out probability
in April that peaks in late May for subadults and in early June for
adults (Figure \@ref(fig:ribbonHOCal); see also 
\@ref(fig:ribbonPredSE)). The behavior was clearly centered
around local solar noon and expanded to other hours later in the spring
as seals entered their molting period. Subadults showed an earlier start
and more intense haul-out activity in April and May. The
young-of-the-year records begin after weaning and the model predictions
demonstrated the development of in-water activities (e.g. swimming, foraging)
in May. Adult females had a more protracted haul-out season compared to
males, and more time was spent hauled out in June compared to adult males and
subadults.

@@

Unlike previous analyses of seal haul-out behavior in spotted and ribbon seals
(e.g. [@verhoef2010a], [@conn2014a]), we also investigated the influence of
sex-age class on haul-out probabilities of both species (but not for bearded seals
because of low sample size). Field identification of age class can be inexact,
particularly when differentiating subadults from adults. In the case of ribbon
seals, subadults often have less distinct ribbons than adults. Spotted seal
pelage cannot be used to reliably discern adults from subadults. Despite these
challenges, we feel the age classifications used in this analysis are useful in
testing for age-related effects on haul-out behavior.

Although ribbon and spotted seals exhibited a
unimodal diel haul-out pattern generally centered around local solar noon, there
were key differences across species, age, and sex that match our understanding
from natural history descriptions of their ecological behavior. Spotted seals
are known to form triads during the breeding season where a female and dependent
pup are accompanied on the ice by a suitor male [@frost2018]. The male waits for
the female to wean the pup and enter estrus, and fends off any other potential
suitor males. Triad formation results in both males and females spending a large
portion of the day hauled out on ice and a protracted spring haul-out season for
both sexes. While females are still nursing the pup and not yet in estrus they
may be less inclined to interrupt their haul out and enter the water where the
suitor male would attempt mating. We see this reflected in the predicted
haul-out patterns, with both males and females exhibiting a broad distribution
of time out of the water throughout the solar day and the season. Ribbon seals
are not known to form triads and our model predicts a progression of increased
haul-out behavior with females starting earlier in the season than males.
Notably, female ribbon seals spend a large portion of the day in the water
during the pupping period, aligning with the hypothesis that ribbon seal females
continue foraging while nursing. Subadult ribbon and spotted seals begin
elevated haul-out behavior earlier in the spring and follow a pattern seen in
other phocids where yearlings and subadults molt first followed by adult females
and males [@reder2003; @thompson1987; @kirkman2003].

We also investigated the influence of weather on haul-out probabilities,
including wind speed, temperature, barometric pressure, precipitation, and wind
chill. These have been investigated for walruses (e.g. Udevitz et al.
[-@udevitz2009]) and a few select studies of ice-associated seals
[@hamilton2018; @perry2017]. In our study, ribbon seals seemed to be the most
influenced by weather, with wind, temperature, and barometric pressure all being
important components of the model. Spotted seals were most affected by wind and
barometric pressure. For bearded seals, the model indicated wind and temperature
had the greatest impact. In general, and as might be expected, seals were more
likely to haul-out when daily temperatures were warmer, winds speeds were lower,
barometric pressure was higher, and precipitation was lower. Those weather
conditions are general indicators of increased solar radiation and lower
convective heat loss, both of which provides energetic benefits (see additional
discussion in Supplemental Material [Exploring Insolation (Solar Radiation) as a
Model Covariate] regarding the potential use of solar radiation directly). Low
winds and precipitation could also enhance predator detection. Our results
highlight the importance of incorporating weather covariates when analyzing
haul-out behavior and calculating availability corrections for aerial surveys.

@@

Consider
the case of employing haul-out probabilities as an aerial survey correction in a
population inhabiting two areas: an ice-free region and a region with sea ice.
Further, assume that half of the population is in each area on average, and that
the probability of hauling out is 1.0 in the ice covered region and 0.0 in the
region without ice.  Denoting the total population as $N$, let us examine what
happens when we use (1) population level availability, and (2) ice-specific
availability as a correction factor.  We'll further assume, for sake of this
example, that detection probability of seals on ice is 1.0, and that the
entirety of the ice covered region is censused. In case (1), our aerial survey
count is $C = 0.5 \cdot N$, with a population-level availability probability of
$\hat{a} = 0.5$. A Horvitz-Thompson-type estimator for abundance is simply 
$\hat{N} = \frac{C}{\hat{a}}$, which has expectation 
$\mathbb{E}(\hat{a}) = \frac{0.5 \cdot N}{0.5} = N$, as desired.  Now
consider case (2).  In this case, the probability of hauling out is 1.0 for
seals in the ice-covered area, so we have 
$\mathbb{E}(\hat{N}) = \frac{0.5 \cdot N}{1.0} = 0.5N$. That
is, we bias abundance because we are not accounting for seals that are away from
ice, and therefore have a zero chance of hauling out. The same logic holds for
any other covariate, such as distance from the ice edge, that has different
values in the surveyed and unsurveyed habitat of the seals (assuming distance
from ice edge is coded with opposite signs inside and outside the pack ice).

@@

Previous attempts to estimate the abundance of phocid seals from aerial survey
data in the Bering and Chukchi seas (e.g. @bengtson2005a, @conn2014a,
@verhoef2014a) have used estimated haul-out probabilities to correct for the
proportion of animals that are in the water and thus unavailable to be counted.
Although several of these studies allowed haul-out probabilities to vary by
day-of-year and time-of-day, they have not accounted for variability among
years, in weather conditions, and in the age-sex class of the sample. In this
paper, we have shown that there can be considerable differences in the number of
seals hauled out on ice based on these factors and subsequent analyses have shown
the potential for considerable bias in abundance estimates if such covariates
are unaccounted for [@conn2023]. We recommend that future
abundance analyses employ availability models that account for them. For
instance, it is relatively straightforward to obtain weather reanalysis products
for times and locations that are surveyed and to construct a relevant correction
factor based on predictions of GLMPMs. The most challenging element in
developing availability correction factors is with annual variability. It can be
difficult to get a sufficient sample size to estimate year-specific correction
factors, particularly because research teams would likely need to tag seals and
conduct aerial surveys concurrently, requiring considerably more personnel
and money. One possible suggestion is to estimate a "shift" parameter within
models for aerial survey counts that allow the peak of haul-out distributions to
be adjusted earlier or later in the year based on the frequency of counts
observed over time. Regardless, researchers should anticipate there being some
unmodeled heterogeneity in availability when using aerial surveys to estimate
Arctic seal abundance. This may require consideration in trend estimation,
as one will not know if moderate differences in abundance estimates are
attributable to changes in abundance or changes in haul-out behavior.

@@

We focused this paper on haul-out behavior of bearded, ribbon, and spotted
seals. Ringed seals are also present in the Bering and Chukchi seas but exhibit
a unique complicating factor. Adult ringed seals build subnivean lairs under the
snow on top of the sea ice, where they haul out and where females rear pups
until conditions are good for basking [@frost2004]. Thus, the wet-dry sensor on
a bio-logger could indicate that an animal is hauled out, but if it is within a
lair, it is not available to be detected during an aerial survey. We hope to
address availability of ringed seals using data from satellite tags, replicate
survey tracks, and auxiliary information about snow depth and timing of melt in
a future study (see, e.g., @lindsay2021).
